Inferring Graphics Programs from Images

### 2017-05-13 paper.tex
section describing how program synthesis works

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2016} with
% \usepackage[nohyperref]{icml2016} above.
\usepackage{amsmath}
\usepackage{hyperref}
\DeclareMathOperator*{\argmin}{arg\,min} % thin space, limits underneath in displays
\DeclareMathOperator{\argmin}{argmin} % no space, limits underneath in displays

@@

High dimensional perceptual input may seem ill matched to the abstract
semantics of a programming language. But programs with constructs like
recursion or iteration produce a simpler \emph{execution trace} of
primitive actions; for our domain the primitive actions are drawing
commands. Our hypothesis is that the execution trace of the program is
better aligned with the perceptual input, and that the trace can act
as a kind of bridge between perception and programs. We test this
hypothesis by developing a model that learns to map from an image to
the execution trace of the graphics program that drew it.  With the
execution trace in hand, we can bring to bear techniques from the
program synthesis community to recover the latent graphics program.
This family of techniques, called \emph{constraint-based program synthesis}~\cite{solar2008program},
work by modeling a set of possible programs inside of a constraint solver,
like a SAT or SMT solver~\cite{de2008z3}.

%% \remark{This is an \emph{excellent} explanation! I would add maybe one
%%   more sentence / citation to elaborate on what you mean by
%%   `techniques from the program synthesis community', as this will be
%%   an unfamiliar concept to some readers, I imagine.}

@@

\section{Neural architecture for inferring image parses}

@@

This network suffices to ``derender'' images like those shown in
Figure~\ref{trainingData}.  We can perform a beam search decoding to
recover what the network thinks is the most likely execution trace for
images like these. But, if the network makes a mistake (predicts an
incorrect line of code), it has no way of recovering from the error.
In order to derender an image with $n$ objects, it must correctly
predict $n$ drawing commands -- so its probability of success will
decrease exponentially in $n$, assuming it has any nonzero chance of
making a mistake.  For added robustness as $n$ becomes large, we treat
the neural network outputs as proposals for a SMC sampling scheme.
For the SMC sampler, we use pixel wise distance as a surrogate for a
likelihood function; see supplement. Figure~\ref{syntheticResults}
compares the neural network with SMC against the neural network by
itself or SMC by itself.  Only the combination of the two passes a
critical test of generalization: when trained on images with $\leq 8$
objects, it successfully parses scenes with many more objects than the
training data.

@@

\section{Synthesizing graphics programs from execution traces}
Although the execution trace of a graphics program describes the parts
of a scene,
it fails to encode higher-level features of the image,
such as repeated motifs,
symmetries or reflections.

\begin{table}[h]
  \begin{tabular}{rl}
  Program$\to$&Command; $\cdots$; Command\\
  Command$\to$&\texttt{circle}(Expression,Expression)\\
  Command$\to$&\texttt{rectangle}(Expression,Expression,Expression,Expression)\\
  Command$\to$&\texttt{LINE}(Expression,Expression,Expression,Expression,Boolean,Boolean)\\
  Command$\to$&\texttt{for}$(0\leq \text{Var}  < \text{Expression})$\texttt{ \{ }Program\texttt{ \}}\\
  Command$\to$&\texttt{REFLECT}$(\text{Axis})$\texttt{ \{ }Program\texttt{ \}}\\
  Expression$\to$&\mathcal{Z}\texttt{ * }Var\texttt{ + }\mathcal{Z}\\
  Var$\to$&A free (unused) variable\\
  \mathcal{Z}$\to$&an integer\\
  Axis$\to$&\texttt{X = }\mathcal{Z}\\
  Axis$\to$&\texttt{Y = }\mathcal{Z}
  \end{tabular}
  \caption{Grammar over graphics programs. We allow loops (\texttt{for}), vertical/horizontal reflections (\texttt{REFLECT}), and affine transformations (\mathcal{Z}\texttt{ * }Var\texttt{ + }\mathcal{Z}).}
  \end{table}

\section{Neural networks for guiding SMC}

Let $L(\cdot | \cdot):\text{image}^2\to \mathcal{R}$ be our likelihood
function: it takes two images, an observed target image and a
hypothesized program output, and gives the likelihood of the observed
image conditioned on the program output. We want to sample from:
\begin{equation}
\probability [p|x]  \propto L(x | \text{render}(p)) \probability [p]
\end{equation}
where $\probability [p]$ is the prior probability of program $p$, and $x$ is the observed image.

Let $p$ be a program with $L$ lines, which we will write as $p = (p_1,p_2,\cdots,p_L)$. Assume the prior factors into:
\begin{equation}
  \probability [p]\propto \prod_{l\leq L}\probability [p_l]
\end{equation}
Define the distribution $q_L(\cdot)$, which happens to be proportional to the above posterior:
\begin{equation}
  q_L(p_1,p_2,\cdots,p_{L - 1},p_L)\propto q_{L - 1}(p_1,p_2,\cdots,p_{L - 1})\times \frac{L(x | \text{render}(p_1,p_2,\cdots,p_{L - 1},p_L))}{L(x | \text{render}(p_1,p_2,\cdots,p_{L - 1}))}\times\probability [p_L]
\end{equation}
Now suppose we have some samples from $q_{L - 1}(\cdot)$, and that we
then sample a $p_L$ from a distribution proportional to $\frac{L(x |
  \text{render}(p_1,p_2,\cdots,p_{L - 1},p_L))}{L(x |
  \text{render}(p_1,p_2,\cdots,p_{L - 1}))}\times\probability [p_L]$.
The resulting programs $p$ are distributed according to $q_L$, and so
are also distributed according to $\probability [p|x]$.

How do we sample $p_L$ from a distribution proportional to $\frac{L(x
  | \text{render}(p_1,p_2,\cdots,p_{L - 1},p_L))}{L(x |
  \text{render}(p_1,p_2,\cdots,p_{L - 1}))}\times\probability [p_L]$?
We have a neural network that takes as input the target image $x$ and
the program so far, and produces a distribution over next lines of
code ($p_L$).  We write $\text{NN}(p_L | p_1,\cdots,p_{L - 1};x)$ for
the distribution output by the neural network. So we can sample from NN and then weight the samples by:
\begin{equation}
  w(p_L) = \frac{\probability [p_L]}{\text{NN}(p_L | p_1,\cdots,p_{L - 1};x)}\times \frac{L(x | \text{render}(p_1,p_2,\cdots,p_{L - 1},p_L))}{L(x | \text{render}(p_1,p_2,\cdots,p_{L - 1}))}
\end{equation}
Then we can resample from these now weighted samples to get a new
population of particles (here programs are particles), where each
program now has $L$ lines instead of $L - 1$.

This procedure can be seen as a particle filter, where each successive
latent variable is another line of code, and the emission
probabilities are successive ratios of likelihoods under $L(\cdot |
\cdot)$.

\textbf{Comments for Dan}. Right now I'm not actually sampling from
the neural network - instead, I enumerate the top few hundred lines of
code suggested by the network, and then weight them by their
likelihoods.
So actually the form of NN is:
\begin{equation}
  \text{NN}(p_L | p_1,\cdots,p_{L - 1};x)\propto \begin{cases}
    1,&\text{ if }p_L \in \text{top hundred neural network proposals}\\
0,&\text{otherwise}.    \end{cases}
\end{equation}
Do you think this is a problem? The neural network puts almost all of its mass on a few guesses.
In order to get the correct line of code I sometimes need to get something like the 50th  top guess,
so I don't want to literally just sample from the distribution suggested by the neural network.

  \begin{algorithm}[tb]
   \caption{Neurally guided SMC}
   \label{guideAlgorithm}
\begin{algorithmic}
  \STATE {\bfseries Input:} Neural network NN, beam size $N$, maximum length $L$, target image $x$
  \STATE {\bfseries Output:} Samples of the program trace
  \STATE Set $B_0 = \{\text{empty program}\}$
  \FOR{$1\leq l\leq L$}
  \FOR{$1\leq n\leq N$}
  \STATE{ $p_n\sim \text{Uniform}(B_{l - 1})$}
  \STATE{ $p'_{n}\sim \text{NN}(\text{render}(p),x)$}
  \STATE{ Define $r_n = p'_n\cdot p_n$}
  \STATE{ Set $\tilde{w}(r_n) = \frac{L(x|r_n)}{L(x|p_n)}\times\frac{\probability [p'_n]}{\probability [p'_n = \text{NN}(\text{render}(p),x)]}$}
  \ENDFOR
  \STATE{ Define $w(p) = \frac{\tilde{w}(p)}{\sum_{p'}\tilde{w}(p')}$}
  \STATE{ Set $B_l$ to be $N$ samples from $r_n$ distributed according to $w(\cdot)$}
  \ENDFOR
  \STATE {\bfseries return} $\{p : p\in B_{l\leq L}, p \text{ is finished}\}$
\end{algorithmic}
  \end{algorithm}

  \pagebreak
  \section{Preliminary extrapolation results} 
\input{extrapolations.tex}

>>>>

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2016} with
% \usepackage[nohyperref]{icml2016} above.
\usepackage{amsmath}
\usepackage{hyperref}
\DeclareMathOperator*{\argmin}{arg\,min} % thin space, limits underneath in displays
\DeclareMathOperator{\argmin}{argmin} % no space, limits underneath in displays
\DeclareMathOperator*{\argmax}{arg\,max} % thin space, limits underneath in displays
\DeclareMathOperator{\argmax}{argmax} % no space, limits underneath in displays

@@

High dimensional perceptual input may seem ill matched to the abstract
semantics of a programming language. But programs with constructs like
recursion or iteration produce a simpler \emph{execution trace} of
primitive actions; for our domain the primitive actions are drawing
commands. Our hypothesis is that the execution trace of the program is
better aligned with the perceptual input, and that the trace can act
as a kind of bridge between perception and programs. We test this
hypothesis by developing a model that learns to map from an image to
the execution trace of the graphics program that drew it.  With the
execution trace in hand, we can bring to bear techniques from the
program synthesis community to recover the latent graphics program.
This family of techniques, called \emph{constraint-based program synthesis}~\cite{solar2008program},
work by modeling a set of possible programs inside of a constraint solver,
like a SAT or SMT solver~\cite{de2008z3}.
These techniques excel at uncovering high-level symbolic structure,
but are not well equipped to deal with real-valued perceptual inputs.

@@

\section{Neural architecture for inferring drawing execution traces}

@@

This network suffices to ``derender'' images like those shown in
Figure~\ref{trainingData}.  We can perform a beam search decoding to
recover what the network thinks is the most likely execution trace for
images like these. But, if the network makes a mistake (predicts an
incorrect line of code), it has no way of recovering from the error.
In order to derender an image with $n$ objects, it must correctly
predict $n$ drawing commands -- so its probability of success will
decrease exponentially in $n$, assuming it has any nonzero chance of
making a mistake.  For added robustness as $n$ becomes large, we treat
the neural network outputs as proposals for a SMC sampling scheme.
For the SMC sampler, we use pixel wise distance as a surrogate for a
likelihood function; see supplement. Figure~\ref{syntheticResults}
compares the neural network with SMC against the neural network by
itself or SMC by itself.  Only the combination of the two passes a
critical test of generalization: when trained on images with $\leq 8$
objects, it successfully parses scenes with many more objects than the
training data.
\begin{wrapfigure}{R}{0.5\textwidth}
  \begin{center}
    \begin{minipage}[t]{2.2cm}\fbox{\includegraphics[width=2cm]{figures/randomScene-58-7.png}}\end{minipage}
    \begin{minipage}[t]{2.2cm}\fbox{\includegraphics[width=2cm]{figures/randomScene-33-5.png}}\end{minipage}
    \begin{minipage}[t]{2.2cm}\fbox{\includegraphics[width=2cm]{figures/randomScene-25-4.png}}\end{minipage}
%    \begin{minipage}[t]{2cm}\includegraphics[width=2cm]{figures/randomScene-26-7.png}\end{minipage}
%    \begin{minipage}[t]{2cm}\includegraphics[width=2cm]{figures/randomScene-19-3.png}\end{minipage}
  \end{center}
  \caption{Network is trained to infer execution traces for figures like the three shown above.}\label{trainingData}
\end{wrapfigure}

@@

\section{Synthesizing graphics programs from execution traces}
Although the execution trace of a graphics program describes the parts
of a scene, it fails to encode higher-level features of the image,
such as repeated motifs, symmetries or reflections.  A \emph{graphics
  program} better describe structures like these,
and we now take as our goal to synthesize simple graphics programs from
their execution traces.

We constrain the space of allowed programs by writing down a context
free grammar over a space of programs. Although it might be desirable
to synthesize programs in a Turing-complete language like Lisp or
Python, a more tractable approach is to specify what in the program
languages community is called a Domain Specific Language (DSL). Our DSL (Table~\ref{DSL})
encodes prior knowledge of what graphics programs tend to look like.

\begin{table}[H]
  \begin{tabular}{rl}
  Program$\to$&Command; $\cdots$; Command\\
  Command$\to$&\texttt{circle}(Expression,Expression)\\
  Command$\to$&\texttt{rectangle}(Expression,Expression,Expression,Expression)\\
  Command$\to$&\texttt{LINE}(Expression,Expression,Expression,Expression,Boolean,Boolean)\\
  Command$\to$&\texttt{for}$(0\leq \text{Var}  < \text{Expression})$\texttt{ \{ }Program\texttt{ \}}\\
  Command$\to$&\texttt{REFLECT}$(\text{Axis})$\texttt{ \{ }Program\texttt{ \}}\\
  Expression$\to$&\mathcal{Z}\texttt{ * }Var\texttt{ + }\mathcal{Z}\\
  Var$\to$&A free (unused) variable\\
  \mathcal{Z}$\to$&an integer\\
  Axis$\to$&\texttt{X = }\mathcal{Z}\\
  Axis$\to$&\texttt{Y = }\mathcal{Z}
  \end{tabular}
  \caption{Grammar over graphics programs. We allow loops (\texttt{for}), vertical/horizontal reflections (\texttt{REFLECT}), and affine transformations (\mathcal{Z}\texttt{ * }Var\texttt{ + }\mathcal{Z}).}\label{DSL}
\end{table}

Given the DSL and a trace $T$, we want a program that evaluates to $T$
and also minimizes some measure of program cost:
\begin{equation}
  \text{program}(T) = \argmin_{\substack{p\in \text{DSL}\\p \text{ evaluates to } T}} \text{cost}(p)\label{programObjective}
\end{equation}
An intuitive measure of program cost is its length.  We define the
cost of a program to be the number of statements it contains, where a
statement is a ``Command'' in Table~\ref{DSL}.

The constrained optimization problem in
equation~\ref{programObjective} is intractable in general, but there
exist efficient-in-practice tools for finding exact solutions to
program synthesis problems like these. We use the state-of-the-art Sketch
tool~\cite{solar2008program}. Describing Sketch's program synthesis
algorithm is beyond the scope of this paper; see supplement.  At a
high level, Sketch takes as input a space of programs, along with a
specification of the program 's behavior and optionally a cost
function.  It translates the synthesis problem into a constraint
satisfaction problem, and then uses a quasiboolean solver to find a
minimum cost program satisfying the specification.  In exchange for
not having any guarantees on how long it will take to find a minimum
cost solution, it comes with the guarantee that it will always find a
globally optimal program.

Why synthesize a graphics program,
if the execution trace already suffices to recover the objects in an image?
Within our domain of hand-drawn figures, graphics program synthesis has several important uses:
\subsection{Extrapolating figures}
Given the source code of a graphics program,
we can automatically tweak the program
to make natural-feeling changes to the figure.
For example,
we can change all of the circles to squares,
were make all of the lines be dashed.
We can also \textbf{extrapolate} figures
by increasing the number of times that loops are executed.
\input{extrapolations.tex}

\subsection{Modeling similarity between figures}

\subsection{Correcting errors made by the neural network}

Let $L(\cdot | \cdot):\text{image}^2\to \mathcal{R}$ be our likelihood
function: it takes two images, an observed target image and a
hypothesized program output, and gives the likelihood of the observed
image conditioned on the program output.
Write $\hat{T}(I)$ for the trace the model predicts for image $I$.

We can extract a few basic features of a program, like its size or how many
loops it has,
and use these features to help predict
whether a trace is the correct explanation for an image.
\begin{equation}
\hat{T}(I) = \argmax_{T} L(I | \text{render}(T)) + \theta\cdot  \phi \left( \text{program}(T) \right)
\end{equation}
where $\phi (\cdot)$ is a feature extractor for programs.
This is equivalent to doing MAP inference in a generative model where the program
is first drawn from a log linear distribution $\propto \exp (\theta\cdot \phi (\text{program}))$,
than the program is executed deterministically,
and then we observe a noisy version of the program's output,
where $L$ is the noise model.

  \pagebreak
  \section{Preliminary extrapolation results}
